Geometrization of characters of tori over non-Archimedean local fields

### 2013-08-15 shanghai.tex
Added comments to the end of the section on the connected case.
Other small changes, too.

\begin{definition}\label{def:GC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\cs{L}\ceq (\gcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(CS.0)}{CS.0} $\gcs{L}$ is a rank-$1$ $\ell$-adic local system on $\bG$;
\labitem{(CS.1)}{CS.1} $\mu: \bm^* \gcs{L} \to \LxL$
  is an isomorphism of sheaves on $\bGxG$ such that the following diagram commutes, 
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
%
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gcs{L} \boxtimes \gcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gcs{L} \boxtimes \bm^* \gcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gcs{L} \tight{0}{\boxtimes}{0} \LxL\\
  \end{tikzcd}
  \]

%
\labitem{(CS.2)}{CS.2} $\phi : \Frob{G}^* \gcs{L} \to \gcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with $\mu$ in the sense that the following diagram commutes.
%
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gcs{L}\boxtimes \Frob{G}^*\gcs{L} \dar{F\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gcs{L} \arrow{rr}{\mu}
    && \LxL\\
  \end{tikzcd}
  \]
\end{enumerate}
In short, a quasicharacter sheaf on $G$  is a rank-$1$ Weil local system $(\gcs{L},\phi)$ on $G$ equipped with an isomorphism $m^* (\gcs{L},\phi) \to (\gcs{L},\phi)\boxtimes (\gcs{L},\phi)$.

@@

@@

With Theorem~\ref{thm:snake} on hand, we can see why neither bounded quasicharacter sheaves nor finite quasicharacter sheaves are up to the task of geometrizing characters of $G(\Fq)$ in the generality we need.  
%
Consider the case when $G$ is the discrete etale group scheme $\ZZ$. Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ and let $\cs{L}$ be a quasicharacter sheaf in the isomorphism class corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. Then $\trFrob{\cs{L}}(1) = \ell$. Suppose $\cs{L}$ was a bounded quasicharacter sheaf. Then the restriction of $\cs{L}$ to the connected component $1\in G$ would be a rank-1 local system on $1$, thus an $\ell$-adic character $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with $\rho_1(\Frob{\Fq}) = \ell$. But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, while $\rho_1(\pi_1(1,{\bar 1}))$ is not compact (it contains $\ell^\ZZ$). Since the Weil sheaf $\gcs{L}$ does not descend to a local system on $G$, the  quasicharacter sheaf$\cs{L}$ is not a bounded quasicharacter sheaf.
%
We can play the same game with the character $\chi : G(\Fq) \to \EEx$ defined by $\chi(1) = 1+\ell$. In this case, the corresponding geometric character $\cs{L}$ is bounded but not finite, since $1+\ell$ is not a root of unity.

These simple examples illustrate why the categories $\QCb(G)$ and $\QCf(G)$ are inadequate to geometrize characters of $G(\Fq)$.

@@

Let $K$ be a non-Archimedean local field with residue field $\Fq$.
Note that we place no restriction on the characteristic of $K$. 
We write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$. 
For any positive integer $d$, set $A_d\ceq \OK/\pK^d$.
Let $T$ be an algebraic torus over $K$.
We write $\Hom_{< d}(T(K),\EEx)$ for the group of characters of $T(K)$ depth less than $d$ and $\Hom_\text{qc}(T(K),\EEx)$ for the group of quasicharacters of $T(K)$.

\section{Greenberg of Neron}
\label{sec:GAN}

\Clifton{Motivate with comments about LLC for tori. Draw triangle. Reduction to GL(1) via induced tori. Comment about Japanese work. Comment about pro-quasi-algebraic. Combine with next section.}

We have already seen that the trace of Frobenius defines a canonical homomorphism $\QCiso{\GN{T}} \to \Hom_\text{ad}(T(K),\EEx)$. 
In this section we show that this homomorphism is functorial, in the following sense.

@@

\begin{definition}
For each positive integer $d$, set $\TT_{d} \ceq \TT \times_\Spec{\OK} \Spec{A_d}$. Observe that $A_d$ is an Artin ring and $\TT_{d}$ is smooth and locally of finite type over $A_d$.
%
Let \cdef{$\GN{T}_{d} = \Gr(\TT_d)$} be the group scheme over $\Fq$ produced by applying
the Greenberg functor to $\TT_{d}$; this is the \cdef{Greenberg transform of $\TT_d$}.
\end{definition}

@@

@@

  Next we show that this isomorphism is functorial.
  When combined with the definition of $\QC(\GN{T})$ and Property~\ref{T(K)}, the first part of Proposition~\ref{prop:pullback} shows that $F_1$ is a functor, while the second part shows that Trace of Frobenius is a natural transformation $t: F_1 \to F_2$. When further combined with Lemma~\ref{prop:product}, we see that $F_1$ is an additive functor and $t: F_1 \to F_2$ is a natural transformation between additive functors.

@@

\bib{SGA1a}{book}{
   author={Grothendieck, Alexander},
   title={Rev\^etements \'etales et groupe fondamental. Fasc. I: Expos\'es 1
   \`a 5},
   series={S\'eminaire de G\'eom\'etrie Alg\'ebrique},
   volume={1960/61},
   publisher={Institut des Hautes \'Etudes Scientifiques},
   place={Paris},
   date={1963},
%   pages={iv+143 pp. (not consecutively paged) (loose errata)},
%   review={\MR{0217087 (36 \#179a)}},
}

>>>>

\begin{definition}\label{def:QC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\cs{L}\ceq (\gcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(CS.0)}{CS.0} $\gcs{L}$ is a rank-$1$ $\ell$-adic local system on $\bG$;
\labitem{(CS.1)}{CS.1} $\mu: \bm^* \gcs{L} \to \LxL$
  is an isomorphism of sheaves on $\bGxG$ such that the following diagram commutes, 
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
%
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gcs{L} \boxtimes \gcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gcs{L} \boxtimes \bm^* \gcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gcs{L} \tight{0}{\boxtimes}{0} \LxL\\
  \end{tikzcd}
  \]

%
\labitem{(CS.2)}{CS.2} $\phi : \Frob{G}^* \gcs{L} \to \gcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with $\mu$ in the sense that the following diagram commutes.
%
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gcs{L}\boxtimes \Frob{G}^*\gcs{L} \dar{F\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gcs{L} \arrow{rr}{\mu}
    && \LxL\\
  \end{tikzcd}
  \]
\end{enumerate}
%In short, a quasicharacter sheaf on $G$  is a rank-$1$ Weil local system $(\gcs{L},\phi)$ on $G$ equipped with an isomorphism $m^* (\gcs{L},\phi) \to (\gcs{L},\phi)\boxtimes (\gcs{L},\phi)$.

@@

As Proposition~\ref{prop:connected} shows, when $G$ is a connected algebraic group over $\Fq$, it is appropriate to replace quasicharacter sheaves on $G$ with the conceptually simpler category $\QCf(G)$ of finite quasicharacter sheaves on $G$. In this context, finite quasicharacter sheaves may also be apprehended as \emph{certain} $\ell$-adic characters of fundamental group $\pi_1(G,{\bar e})$. As explained in \cite{Kamgarpour}*{\S~2}, the relevant characters of $\pi_1(G,{\bar e})$ are precisely those that factor through the quotient of $\pi_1(G,{\bar e})$ denoted there by $\Pi_\text{disc}(G)$.
%\[\pi_1(G,{\bar e}) \to \Pi_\text{et}(G) \to \Pi_\text{et}(G)_{\Gal(\bFq/\Fq)} = \Pi_\text{disc}(G)\]
From this is is apparent that the category of discrete isogenies to $G$ forms a Galois category, in the sense of \cite{SGA1}*{Expos\'e V, \S 4}. Moreover, since $\Pi_\text{disc}(G) \iso G(\Fq)$, as shown in \cite{Kamgarpour}*{App'x B}, this provides an alternate proof of $\QCiso{G} \to G(\Fq)^*$.

In principle, something similar is possible in the more general context of this paper, where $G$ is a smooth group scheme lft over $\Fq$, to which we now return. Every quasicharacter sheaf $(\gcs{L},\mu,\phi)$ on $G$ determines a Weil sheaf $(\gcs{L},\phi)$ on $G$. So, if we choose a geometric point ${\bar g}_x$ in each component $\bG^x$ of $G$, then this choice can be used to convert the Weil sheaf $(\gcs{L},\phi)$ into an $\ell$-adic character of $\prod_{x\in \pi_0(\bG)}\W(\bG^x, {\bar g}_x)$. It proved cumbersome, however, to manage various choices of families of base points and track the action of $\Gal(\bFq/\Fq)$ on this (generally) infinite product of Weil groups, and even more difficult to identity the relevant quotient of this group. In this light, Definition~\ref{def:QC} seemed simple.

@@

With Theorem~\ref{thm:snake} on hand, 
we can see why neither bounded quasicharacter sheaves 
nor finite quasicharacter sheaves are up to the task of 
geometrizing characters of $G(\Fq)$ in the generality we need.  
%
Consider the case when $G$ is the discrete etale group scheme $\ZZ$. 
Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ 
and let $\cs{L}$ be a quasicharacter sheaf in the isomorphism class 
corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. 
Then $\trFrob{\cs{L}}(1) = \ell$. Suppose $\cs{L}$ was a bounded quasicharacter sheaf. 
Then the restriction of $\cs{L}$ to the connected component $1\in G$ 
would be a rank-1 $\ell$-adic local system on $1$, 
thus an $\ell$-adic character $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with $\rho_1(\Frob{\Fq}) = \ell$. But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, 
while $\rho_1(\pi_1(1,{\bar 1}))$ is not compact in $\EEx$ (it contains $\ell^\ZZ$). 
So the Weil sheaf $\gcs{L}$ does not descend to a local system on $G$.
It follows that the quasicharacter sheaf $\cs{L}$ is not a bounded quasicharacter sheaf.
%
We can play the same game with the character $\chi : G(\Fq) \to \EEx$ 
defined by $\chi(1) = 1+\ell$. In this case, the corresponding quasicharacter sheaf $\cs{L}$ is bounded but not finite, since $1+\ell$ is not a root of unity. 
%
These simple examples illustrate why the categories $\QCb(G)$ and $\QCf(G)$ are inadequate to geometrize characters of $G(\Fq)$.

@@

Let $K$ be a non-Archimedean local field with residue field $\Fq$.
We place no restriction on the characteristic of $K$. 
We write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$. 
For any positive integer $d$, set $A_d\ceq \OK/\pK^d$.
Let $T$ be an algebraic torus over $K$.
We write $\Hom_{< d}(T(K),\EEx)$ for the group of characters of $T(K)$ depth less than $d$ 
and $\Hom_{<\infty}(T(K),\EEx)$ for the group of characters of $T(K)$ of finite-depth.

\section{Greenberg of Neron}
\label{sec:GN}

@@

\begin{definition}
For each positive integer $d$, set $\TT_{d} \ceq \TT \times_\Spec{\OK} \Spec{A_d}$. Observe that $A_d$ is an Artin ring and $\TT_{d}$ is smooth and locally of finite type over $A_d$.
%
Let \cdef{$\GN{T}_{d} = \Gr(\TT_d)$} be the group scheme over $\Fq$ produced by applying
the Greenberg functor to $\TT_{d}$; this is the \emph{Greenberg transform of $\TT_d$}.
\end{definition}

@@

\Clifton{Motivate with comments about LLC for tori. Draw triangle. Reduction to GL(1) via induced tori. Comment about Japanese work. Comment about pro-quasi-algebraic. }

@@

  Next we show that this isomorphism is functorial.
The first part of Proposition~\ref{prop:pullback} shows that $F_1$ is a functor, while the second part shows that Trace of Frobenius is a natural transformation $t: F_1 \to F_2$. When further combined with Lemma~\ref{prop:product}, we see that $F_1$ is an additive functor and $t: F_1 \to F_2$ is a natural transformation between additive functors.

@@

\bib{SGA1}{collection}{
   title={Rev\^etements \'etales et groupe fondamental (SGA 1)},
%   language={French},
   series={Documents Math\'ematiques (Paris), 3},
   note={S\'eminaire de g\'eom\'etrie alg\'ebrique du Bois Marie 1960--61.
   [Algebraic Geometry Seminar of Bois Marie 1960-61];
   Directed by A. Grothendieck;
   With two papers by M. Raynaud;
   Updated and annotated reprint of the 1971 original [Lecture Notes in
   Math., 224, Springer, Berlin;  MR0354651 (50 \#7129)]},
   publisher={Soci\'et\'e Math\'ematique de France},
   place={Paris},
   date={2003},
   pages={xviii+327},
%  isbn={2-85629-141-4},
%   review={\MR{2017446 (2004g:14017)}},
}
